---
title: "Biscuiteer User Guide"
date: "`r BiocStyle::doc_date()`"
package: "`r BiocStyle::pkg_ver('biscuiteer')`"
output:
  BiocStyle::html_document:
    highlight: pygments
    toc_float: true
    fig_width: 8
    fig_height: 6
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Biscuiteer User Guide}
  %\VignetteEncoding[utf8]{inputenc}
---

# Biscuiteer

`biscuiteer` is package to process output from
[biscuit](https://github.com/zwdzwd/biscuit) into
[bsseq](https://bioconductor.org/packages/bsseq) objects. It includes a number
of features, such as VCF header parsing, shrunken M-value calculations (which
can be used for compartment inference), and age inference are included. However,
the task of locus- and region-level differential methylation inference is
delegated to other packages (such as `dmrseq`).

# Quick Start

## Installing

From Bioconductor,
```{r, eval=FALSE}
if (!requireNamespace("BiocManager", quietly=TRUE))
    install.packages("BiocManager")
BiocManager::install("biscuiteer")
```

A development version is available on GitHub and can be installed via:
```{r, eval=FALSE}
if (!requireNamespace("BiocManager", quietly=TRUE))
    install.packages("BiocManager")
BiocManager::install("trichelab/biscuiteer")
```

## Loading Data

`biscuiteer` can load either headered of header-free BED files produced from
`biscuit vcf2bed` or `biscuit mergecg`. In either case, a VCF file is needed
when loading `biscuit` output. For practical purposes, only the VCF header is
for `biscuiteer`. However, it is encouraged that the user keep the entire VCF,
as `biscuit` can be used to call SNVs and allows for structural variant
detection in a similar manner to typical whole-genome sequencing tools.
Furthermore, `biscuit` records the version of the software and the calling
arguments used during processing the output VCF, which allows for better
reproducibility.

NOTE: Both the input BED and VCF files must be tabix'ed before being input to
`biscuiteer`. This can be done by running `bgzip biscuit_output.xxx` followed by
`tabix -p xxx biscuit_output.xxx.gz`, where `xxx` is either `bed` or `vcf`.

Data can be loaded using the `read.biscuit` function in `biscuiteer`:
```{r}
library(biscuiteer)

tcga_bed <- system.file("extdata", "TCGA_BLCA_A13J_chr11p15_merged.bed.gz",
                        package = "biscuiteer")
tcga_vcf <- system.file("extdata", "TCGA_BLCA_A13J_header_only.vcf.gz",
                        package = "biscuiteer")
bisc <- read.biscuit(BEDfile = tcga_bed, VCFfile = tcga_vcf,
                     merged = TRUE, genome = "hg38")
```

Metadata from the `biscuit` output can be viewed via:
```{r}
biscuitMetadata(bisc)
```

If further information about the VCF header is desired,
```{r}
metadata(bisc)$vcfHeader
```

## Combining Results

In the instance where you have two separate BED files that you would like to
analyze in a single bsseq object, you can combine the files using `unionize`,
which is a wrapper around the bsseq function, `combine`.
```{r}
tcga_mrg <- system.file("extdata",
                        "TCGA_BLCA_A13J_chr11p15_unmerged.bed.gz",
                        package = "biscuiteer")
tcga_shf <- system.file("extdata",
                        "TCGA_BLCA_A13J_chr11p15_shuffled_unmerged.bed.gz",
                         package = "biscuiteer")
tcga_mvcf <- system.file("extdata",
                         "TCGA_BLCA_A13J_header_only.vcf.gz",
                         package = "biscuiteer")
tcga_svcf <- system.file("extdata",
                         "TCGA_BLCA_A13J_shuffled_header_only.vcf.gz",
                         package = "biscuiteer")
bisc1 <- read.biscuit(BEDfile = tcga_mrg, VCFfile = tcga_mvcf,
                      merged = FALSE, genome = "hg38")
bisc2 <- read.biscuit(BEDfile = tcga_shf, VCFfile = tcga_svcf,
                      merged = FALSE, genome = "hg38")

comb <- unionize(bisc1, bisc2)
```

# Analysis Functionality

A handful of analysis paths are available in `biscuiteer`, including A/B
comparment inference, age estimation from WGBS data, hypermethylation of
Polycomb Repressor Complex (PRC) binding sites, and hypomethylation of
CpG-poor "partially methylated domains" (PMDs).

## Inputs for A/B Compartment Inference

When performing A/B compartment inference, the goal is to have something that
has roughly gaussian error. `getLogitFracMeth` uses Dirichlet smoothing to turn
raw measurements into lightly moderated, logit-transformed methylated-fraction
estimates, which can the be used as inputs to
[compartmap](https://bioconductor.org/packages/release/bioc/html/compartmap.html)

```{r}
tcga_bed <- system.file("extdata", "TCGA_BLCA_A13J_chr11p15_merged.bed.gz",
                        package = "biscuiteer")
tcga_vcf <- system.file("extdata", "TCGA_BLCA_A13J_header_only.vcf.gz",
                        package = "biscuiteer")
bisc <- read.biscuit(BEDfile = tcga_bed, VCFfile = tcga_vcf,
                     merged = TRUE, genome = "hg38")

reg <- GRanges(seqnames = rep("chr11",5),
               strand = rep("*",5),
               ranges = IRanges(start = c(0,2.8e6,1.17e7,1.38e7,1.69e7),
                                end= c(2.8e6,1.17e7,1.38e7,1.69e7,2.2e7))
              )

frac <- getLogitFracMeth(bisc, minSamp = 1, r = reg)
frac
```


## Age Estimation

`biscuiteer` has the functionalitity to guess the age of the sample(s) provided
using the Horvath-style "clock" models (see
[Horvath, 2013](https://genomebiology.biomedcentral.com/articles/10.1186/gb-2013-14-10-r115)
for more information).

NOTE: The prediction accuracy of this function is entirely dependent on the
parameters set by the user. As such, the defaults (as shown in the example
below) should only be used as a starting point for exploration by the user.

NOTE: Please cite the appropriate papers for the epigenetic "clock" chosen:
* For `horvath` or `horvathshrunk`
    * Horvath, Genome Biology, 2013
* For `hannum`
    * Hannum et al., Molecular Cell, 2013
* For `skinandblood`
    * Horvath et al., Aging, 2018

```{r}
tcga_mrg <- system.file("extdata",
                        "TCGA_BLCA_A13J_chr11p15_unmerged.bed.gz",
                        package = "biscuiteer")
tcga_shf <- system.file("extdata",
                        "TCGA_BLCA_A13J_chr11p15_shuffled_unmerged.bed.gz",
                         package = "biscuiteer")
tcga_mvcf <- system.file("extdata",
                         "TCGA_BLCA_A13J_header_only.vcf.gz",
                         package = "biscuiteer")
tcga_svcf <- system.file("extdata",
                         "TCGA_BLCA_A13J_shuffled_header_only.vcf.gz",
                         package = "biscuiteer")
bisc1 <- read.biscuit(BEDfile = tcga_mrg, VCFfile = tcga_mvcf,
                      merged = FALSE, genome = "hg38")
bisc2 <- read.biscuit(BEDfile = tcga_shf, VCFfile = tcga_svcf,
                      merged = FALSE, genome = "hg38")

comb <- unionize(bisc1, bisc2)
ages <- WGBSage(comb, "horvath")
ages
```

## Hypermethylation of PRCs and Hypomethylation of PMDs

To generate a shorthand summary of the hypermethylation of PRCs and
hypomethylation of PMDs, the `CpGindex` function in `biscuiteer` can be used.

```{r}
bisc.CpGindex <- CpGindex(bisc)
show(bisc.CpGindex)
bisc.CpGindex@hyperMethRegions
```
